---
title: "Weather and Corn Yield Regressions"
author: "Tad Trimarco"
date: "2/28/2022"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab)
library(rnassqs)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}
# daily max temperature
# dimensions: counties x days x years
prism <- readMat("prismiowa.mat")
# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1]
t_1981_c1[366]
plot(1:366, t_1981_c1, type = "l")
ggplot() +
  geom_line(mapping = aes(x=1:366, y = t_1981_c1)) +
  theme_bw() +
  xlab("day of year") +
  ylab("daily maximum temperature (°C)") +
  ggtitle("Daily Maximum Temperature, Iowa County #1")
```
```{r tidying up}
# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years)
# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)
# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax")
tmaxdf <- tibble(tmaxdf)
```

## Temperature trends

### Summer temperature trends: Winneshiek County

```{r temp trends}
tmaxdf$doy <- as.numeric(tmaxdf$doy)
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))
winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))
ggplot(winnesummer, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)
lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax)
```

### Winter Temperatures - Winneshiek County

```{r winter temps}
winnewinter <- tmaxdf %>%
  filter(countyfp==191 & (doy <= 59 | doy >= 335) & !is.na(tmax)) %>%
  group_by(year) %>%
  summarize(meantmax = mean(tmax))
ggplot(winnewinter, mapping = aes(x = year, y = meantmax)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Tmax (°C)") +
  geom_smooth(method = lm)
lm_wintertmax <- lm(meantmax ~ year, winnewinter)
summary(lm_wintertmax)
```

### Multiple regression -- Quadratic time trend

```{r quadratic temp trend}
winnewinter$yearsq <- winnewinter$year^2
lm_wintertmaxquad <- lm(meantmax ~ year + yearsq, winnewinter)
summary(lm_wintertmaxquad)
winnewinter$fitted <- lm_wintertmaxquad$fitted.values
ggplot(winnewinter) +
  geom_point(mapping = aes(x = year, y = meantmax)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "tmax")
```

### Download NASS corn yield data

```{r yield download}
# set our API key with NASS
nassqs_auth(key = "C8B5375C-D03A-3E4D-AF9E-90078918CDBA")
# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")
# download
cornyieldsall <- nassqs_yields(params)
cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)
# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)
```

## Assignment

### Question 1a: Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?

```{r}
cornyields_winn <- cornyields %>%
  filter(county_name == "WINNESHIEK")

lm_cornyields_winn <- lm(yield ~ year, cornyields_winn)
summary(lm_cornyields_winn)

ggplot(cornyields_winn, mapping = aes(x = year, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "year", y = "Yield (bu/ac)") +
  geom_smooth(method = lm)
```

There is absolutely a strong linear trend between yield and year. It appears that as time progresses, yield increases near linearly. 


### Question 1b: Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 

```{r}
cornyields_winn$yearsq <- cornyields_winn$year^2
lm_cornyieldsquad <- lm(yield ~ year + yearsq, cornyields_winn)
summary(lm_cornyieldsquad)
cornyields_winn$fitted <- lm_cornyieldsquad$fitted.values
ggplot(cornyields_winn) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "Yield (bu/ac)")
```

While a quadratic trend certainly exists between yield and year, the slight upward bend to the fitted curve indicates that there is not evidence for slowing yield growth. If anything, yield is growing slightly faster, though not to a level that is apparently different than a linear fit. It seems as though a linear fit for yield growth is as appropriate as a quadratic fit over this time period. 

### Question 2 -- Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.

```{r}
yield_temp_summer <- inner_join(winnesummer, cornyields_winn, by = "year") %>%
  mutate(Tmaxsq = (meantmax^2))

lm_temp_yield <- lm(yield ~ meantmax, yield_temp_summer)
summary(lm_temp_yield)
yield_temp_summer$fitted_3 <- lm_temp_yield$fitted.values
ggplot(yield_temp_summer) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted_3)) +
  theme_bw() + 
  labs(x = "year", y = "yield", title = "Tmax Fitted")

lm_temp_year_yield <- lm(yield ~ year + meantmax, yield_temp_summer)
summary(lm_temp_year_yield)
yield_temp_summer$fitted <- lm_temp_year_yield$fitted.values
ggplot(yield_temp_summer) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted)) +
  theme_bw() +
  labs(x = "year", y = "yield", title = "Tmax and Year Fitted")

lm_temp_tmax2_yield <- lm(yield ~ Tmaxsq + meantmax, yield_temp_summer)
summary(lm_temp_tmax2_yield)
yield_temp_summer$fitted_2 <- lm_temp_tmax2_yield$fitted.values
ggplot(yield_temp_summer) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = fitted_2)) +
  theme_bw() +
  labs(x = "year", y = "yield", title = "Tmax^2 and Tmax Fitted")


```

Of the three plots generated, the model that incorporates year and Tmax seems to be the best fitting. Maximum summer temperatures seem to largely be unrelated to yield. This is supported by our model summary, where Tmax and yield fitted produces a p-value of 0.29. However, when year is included in the model, the p-value decreases to 3.76e-11, indicating a very good fit. The model that fits both Tmax and Tmaxs-squared has a p-value < 0.05, indicating that it does somewhat fit the data, but nearly as well as the year + meantmax model. 


### Question 3 -- Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.

```{r}
cornyields$countyfp <- as.factor(cornyields$county_ansi)

tmax_all <- tmaxdf %>%
  filter(year == 2018) %>%
  select(countyfp, tmax) %>%
  group_by(countyfp) %>%
  summarize(meantmax = mean(tmax, na.rm = TRUE))

cornyields_2018 <- cornyields %>%
  filter(year == 2018)

all_county <- inner_join(cornyields_2018, tmax_all, by = "countyfp") %>%
  mutate(Tmaxsq = (meantmax^2))

lm_cornyields_2018<- lm(yield ~ meantmax + Tmaxsq, all_county)
summary(lm_cornyields_2018)
all_county$fitted <- lm_cornyields_2018$fitted.values

ggplot(all_county, mapping = aes(x = meantmax, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Mean Tmax", y = "Yield") +
  geom_smooth(method = lm) + 
  geom_line(mapping = aes(x = meantmax, y = fitted))

```

It generally appears that a model fitting Tmax to yield across all states in 2018 is a poor estimate of yield. However, a quadratic equation of Tmaxsq and meantmax fits the model nicely, with a p-value of 1.46e-05. 


### Question 4 -- Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.

```{r}

all_county_years <- inner_join(cornyields, tmax_all, by = "countyfp") %>%
  mutate(Tmaxsq = (meantmax^2))

lm_cornyields_2018_4<- lm(yield ~ meantmax + Tmaxsq + countyfp, all_county_years)
summary(lm_cornyields_2018_4)
all_county_years$fitted_4 <- lm_cornyields_2018_4$fitted.values

ggplot(all_county_years, mapping = aes(x = fitted_4, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Fitted Yield", y = "Actual Yield") + 
  geom_smooth(method = lm)

```


It seems that fitted yield and actual yield are related, but the spread of residuals is a little concerning. It seems like the significance of my temperature coefficiencts change a lot, and now none of them have a p-value < 0.05 (meantmax = 0.165, Tmaxsq = 0.142 ). Overall, the fitted model does have a positive linear relationship with the actual yield in each county, but the spread and residuals are so great that I am a little doubtful of the model's effectiveness. The p-value for model is 2.2e-16, but I feel that this is because there are so many degrees of freedom, rather than an accurate evaluation of the model. We may may our p-value too small by choosing too many variables. This is somewhat supported by the R-squared value of 0.082. I would feel a lot more comfortable making conclusions using one of my earlier models, particularly because they showed that year has a greater impact on yield than summer temperatures. However, I do think this model does a pretty decent job at estimating yield.



### Question 5 -- Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.

```{r}

# set our API key with NASS
nassqs_auth(key = "C8B5375C-D03A-3E4D-AF9E-90078918CDBA")
# parameters to query on 
params_s <- list(commodity_desc = "SOYBEANS", statisticcat_desc = "YIELD", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")
# download
beanyieldsall <- nassqs_yields(params_s)
beanyieldsall$county_ansi <- as.numeric(beanyieldsall$county_ansi)
beanyieldsall$yield <- as.numeric(beanyieldsall$Value)
# clean and filter this dataset
beanyields <- select(beanyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
beanyields <- tibble(beanyields)

beanyields$countyfp <- as.factor(beanyields$county_ansi)

all_county_years_beans <- inner_join(beanyields, tmax_all, by = "countyfp") %>%
  mutate(Tmaxsq = (meantmax^2))

lm_beanyields<- lm(yield ~ meantmax + Tmaxsq + countyfp, all_county_years_beans)
summary(lm_beanyields)
all_county_years_beans$fitted <- lm_beanyields$fitted.values

ggplot(all_county_years_beans, mapping = aes(x = fitted, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Fitted Yield", y = "Actual Yield") + 
  geom_smooth(method = lm)


```

It seems that yet again, using our panelled method, we produce a good p-value of 2.2e-16 in this model. Our theoretical yield is closely related to our actual yield in this model, though again I am somewhat concerned by the number of degrees of freedom. It still feels like we're falsely creating a low p-value by choosing too many variables for our model. Still, it is promising to see that our model works well, using the same Tmax and Tmax^2 from our last panelled model. 

### Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.

### Bonus #2: Challenge question - map trends in corn yields by county across Iowa. Interpret your map.